---
title: "Introduction to R"
subtitle: "Bonus Material"
author: ""
filters:
  - diagram
diagram:
  engine:
    tikz:
      header-includes:
        - '\usepackage{adjustbox,xcolor}'
        - '\usetikzlibrary{arrows, shapes}'
        - '\definecolor{UMdblue}{RGB}{0,28,61}'
        - '\definecolor{UMlblue}{RGB}{0,162,219}'
        - '\definecolor{UMorangered}{RGB}{232,78,16}'
        - '\definecolor{UMorange}{RGB}{243,148,37}'
        - '\definecolor{UMred}{RGB}{174,11,18}'
---

# Overview

1.  [Bonus Material Session 3](#Session3)


# Bonus Material for Session 3 {Session3}

## Exercise 3.3: Create your own function

1.  Load the `courses` data frame, which contains information about several courses at a school of business and economics. If the file is located in a folder called "data" within your working directory, you can load it as

```{r echo=TRUE}
load("data/courses.RData")
```

2.  Create a function that takes as input the `courses` data frame. The function output should be the total number of courses that use student tutors.

- There are many ways to count the number of courses that require student tutors. Probably the easiest is to directly do calculations with the logical values: `TRUE` is treated as 1, `FALSE` as 0:

```{r}
c(TRUE + FALSE, TRUE + TRUE, TRUE * FALSE)
```

## Exercise 3.3: Create your own function

2.  We now want to extend the function to give a second output. This output should contain a data frame with two columns: the first is the course code, the second is the number of tutorials for each course.

- The number of tutorials should be calculated as the number of students divided by the number of students per tutorial group, and then rounded up.
- The number of students per tutorial group should be a second argument of the function, with a default value of 15.
- To round up, we can use the function `ceiling()`.


# Bonus: Flow Control {#flowcontrol}

## Flow Control

- Often, expressions (calculations, estimations, simulations, plots...) should only be executed under certain conditions and/or repeated multiple times.

- In such cases, we need:

  + Conditional execution (`if`, `else`)
  
  + Loops (`for`, `while`)

## Conditional Execution

- The general syntax is as follows:
```{r ifelse, eval = FALSE}
if ( expr ) {
  ## some code evaluated if expr == TRUE
} else {
  ## some other code
}
```

- A simple example:
```{r ifelse_example}
x <- 5
if (x > 0) {
  y <- 1       ## if x is positive, y should get value 1
} else {
  y <- -1
}
print(y)
```

## Conditional Execution

- Be careful with vector-valued expressions:

```{r ifelse_vector, error=TRUE}
y <- c(5, 3, 2)

if (y > 4) {
	print("All elements of y a larger than 4")
}
```

If you want to check whether all elements of a vector satisfy a certain property, the quantifiers `any()` and `all()` are useful:

```{r ifelse_quantor}
if (all(y > 4) ) {        ## all() is "TRUE" if the argument is a vector of all TRUE values
    print("All y values are greater than 4")
} else {
    print("At least one element of y is less than or equal to 4!")
}
```

## Conditional Execution with Multipe Options

- There may be more than two cases to consider, for example:
$$y = \cases{0 \text{ for  } x \leq 0\\
                4 \text{ for  } 0< x \leq 5\\
                6 \text{ otherwise.}}$$

```{r ifelse_more, eval = FALSE}
if ( x <= 0 ) {
  y <- 0
} else if (x <= 5 ) {
  y <- 4
} else {
  y <- 6
}
```

## Conditional Execution per Element

- Sometimes we want to check vectors and return a separate value for each entry. For this, we use the `ifelse()` function:
```{r ifelse_vectorized}
x <- c(3, NA, 2)
## is.na() checks if the i put is a missing value
ifelse( is.na(x), "Missing", "Not Missing")

## an elaborate way to calculate absolute values!
x <- c(-3, 5, -8, 2)
ifelse( x < 0, -x, x)
```

## Exercise 3.8: Conditional Execution

- Write a function `if_test` that takes two objects, `x` and `y`, and checks whether `x` is numeric and `y` is of type character.

- If both conditions are met, the function should print "Super"; otherwise, it should indicate which of the two objects `x` or `y` does not meet the required condition.

- Test your function with:
```{r ex_5.1_test, eval = FALSE}
if_test(5, "char")

if_test("abc", 2)

if_test(list(1,2), matrix("a", 3, 3))
```

- Hint: Use `is.character()` and `is.numeric()`.

## Loops

Frequently, small code blocks need to be executed multiple times. Loops help with this. In R, there are multiple types of loops:

- `while( Condition) { Code }`: The code `Code` executes as long as `Condition` is `TRUE`.

- `for (counter in values) { Code }`: The code `Code` runs once for each element in the vector `values`. The variable `counter` takes on the value of each element during each iteration.


## Loops: Examples

- Some examples
```{r example_loops}
vec <- c("One", "Two", "Three")
for (v in vec) {
	print(v)
}

for (i in 1:5) {
  print(i + 2)
}
```

## Combining for and if

- The combination of loops and conditional execution makes for a simple yet very powerful programming interface.

- The major advantage is that it allows to automate certain repetitive operations.

- This code checks for every column of a matrix (or data frame) `dataset` whether it is skewed to the right, and if so, applies a log transformation. (Which statistically is not the brightest idea...)

- We use the `moments` package for calculating the skewness.
```{r eval=FALSE}
library(moments)
n_col <- ncol(dataset)      # check how many columns there are
transformed_data <- dataset # copy the data into a new matrix holding the transformed data
for (i in 1:n_col) {        # loop over all columns
	skew <- skewness(dataset[, i])  # compute the skewness
	if (skew > 1) {
		transformed_data[, i] <- log(dataset[, i]) # log-transform the data
	}
}
```

## Exercise 3.9: Impute Missing Values

- Write a function that checks for every element of a vector whether it is a missing value.  If a missing value is found, it is to be imputed as the average of the value before it and the value after it.

- If the first value in the vector is missing, it should get the value of the second element. If the last value is missing, it should get the value of the preceding element.

- To keep things simple, we assume that there are never two consecutive missing values, such that only the rules above have to be programmed.

- Although there are many ways to create such a function, we focus on using a `for` loop and `if` statements.

- Test your function on the following vector:
```{r eval=FALSE}
c(NA, 2, 3, NA, 5, NA)
```

